---
title: "Data Preparation for SNARX and NARX"
author: "Michael Demone Jackson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Data Preparation:  narxdataprep

## Overview
________________________________________________________________________________

**CoFESNARXdata** and **CoFESSNARXdata** prepare your data for the CoFES NARX 
and CoFES SNARX networks respectively.  **CoFESNARXdata** returns the the number
of target lags ($d_y$) and exegenous lags  ($d_x$) that you request in order to
run the NARX model that you desire while the **CoFESSNARXdata** returns the same
but also adds a seasonal lag (*s*) for the target series.  Both functions allows
users to set the number of lags in the future that the target series via the 
(*delay*=1) parameter. 


## Utility Functions
There are two utility functions: (**returnTARGET(data)**,**returnX(data,T)**)

- **returnTARGET(data)** takes the data frame created by 
**CoFESNARXdata** or **CoFESSNARXdata** and returns the target series as a dataframe.

- **returnX(data,T)**takes the data frame created by**CoFESNARXdata** or 
**CoFESSNARXdata** and returns the regressors in a dataframe for your respective network.  

**Narx**: $$y[t+delay]=f(x[t-d_x],...,x[t-1],x[t], y[t-d_y],...,y[t-l], \Theta)$$

- ($d_y$) The number of passed values/lags of $y_t$ used to predict $y_{(t+delay)}$

- ($d_x$) The number of passed values/lags of $x_t$ used to predict $y_{(t+delay)}$

- (*delay*=1) The number of lags in the future that the networks predicts.  ex) $y_{(t+3)}$

**SNarx**:  
$$y[t+delay]=f(x[t-d_x],...,x[t-1],x[t],\underline{y[t-s]},y[t-d_y],...,y[t-l], \Theta)$$

- (*s*):  Seasonal Lag of $y_t$ used to predict $y_{(t+delay)}$

## CoFESNARXdata Exmple (NARX data prep)

1)  With Date

```{r warning=FALSE}

dir.create("inst/doc")
file.copy(dir("vignettes", full.names=TRUE), "inst/doc", overwrite=TRUE)

library(CoFESSNARX)
library(dplyr)

set.seed(123)
N <- 450 
n <- seq(1:N)
a <- n/10+4*sin(n/10)+sample(-1:6,N,replace=T)+rnorm(N)
b <- n/8+4*cos(n/10)+sample(-3:3,N,replace=T)+rnorm(N) 
c <- n/6+4*sin(n/10)+sample(-5:1,N,replace=T)+rnorm(N)
y <- (a+b+c)/3+rnorm(N)


date<-seq(as.Date("2018-1-1"), as.Date("2019-3-26"), by="days")

x <- data.frame(date,a,b,c)  
y <- data.frame(date, y)

dx <- 2 # The AR order for the Exegenous series 
dy <- 2 # The AR order for the Target series 
delay <- 10 # The number of time steps that the target is in the future
date <- T # Date or no Date Boolean

data<-CoFESSNARX::CoFESNARXdata(y,x,dy,dx,delay,date)

head(data)


y <- CoFESSNARX::returnTARGET(data)
x <- CoFESSNARX::returnX(data,T)
print(head(y))
print(head(x))
```


1)  Without date

```{r warning=FALSE}

library(CoFESSNARX)
library(dplyr)
set.seed(123)
N <- 450 
n <- seq(1:N)
a <- n/10+4*sin(n/10)+sample(-1:6,N,replace=T)+rnorm(N)
b <- n/8+4*cos(n/10)+sample(-3:3,N,replace=T)+rnorm(N) 
c <- n/6+4*sin(n/10)+sample(-5:1,N,replace=T)+rnorm(N)
y <- (a+b+c)/3+rnorm(N)


x <- data.frame(a,b,c)
y <- data.frame(y)


dx <- 2 # The AR order for the Exegenous series 
dy <- 2 # The AR order for the Target series 
delay <- 10 # The number of time steps that the target is in the future
date <- F # Date or no Date Boolean

data<-CoFESSNARX::CoFESNARXdata(y,x,dy,dx,delay,date)

head(data)


y <- CoFESSNARX::returnTARGET(data)
x <- CoFESSNARX::returnX(data,F)
print(head(y))
print(head(x))
```


## CoFESSNARXdata Exmple  (SNARX data prep)

1)  With Date

```{r warning=FALSE}

library(CoFESSNARX)
library(dplyr)

set.seed(123)
N <- 450 
n <- seq(1:N)
a <- n/10+4*sin(n/10)+sample(-1:6,N,replace=T)+rnorm(N)
b <- n/8+4*cos(n/10)+sample(-3:3,N,replace=T)+rnorm(N) 
c <- n/6+4*sin(n/10)+sample(-5:1,N,replace=T)+rnorm(N)
y <- (a+b+c)/3+rnorm(N)


date<-seq(as.Date("2018-1-1"), as.Date("2019-3-26"), by="days")

x <- data.frame(date,a,b,c)
y <- data.frame(date, y)


dx <- 2 # The AR order for the Exegenous series 
dy <- 2 # The AR order for the Target series 
delay <- 10 # The number of time steps that the target is in the future
date <- T # Date or no Date Boolean
s <- 12

data<-CoFESSNARX::CoFESSNARXdata(y,x,dy,dx,delay,s,date)

head(data)


y <- CoFESSNARX::returnTARGET(data)
x <- CoFESSNARX::returnX(data,T)
print(head(y))
print(head(x))
```


1)  Without date

```{r warning=FALSE}

library(CoFESSNARX)
library(dplyr)

set.seed(123)
N <- 450 
n <- seq(1:N)
a <- n/10+4*sin(n/10)+sample(-1:6,N,replace=T)+rnorm(N)
b <- n/8+4*cos(n/10)+sample(-3:3,N,replace=T)+rnorm(N) 
c <- n/6+4*sin(n/10)+sample(-5:1,N,replace=T)+rnorm(N)
y <- (a+b+c)/3+rnorm(N)


x <- data.frame(a,b,c)
y <- data.frame(y)


dx <- 2 # The AR order for the Exegenous series 
dy <- 2 # The AR order for the Target series 
delay <- 10 # The number of time steps that the target is in the future
date <- F # Date or no Date Boolean
s <- 12
data<-CoFESSNARX::CoFESSNARXdata(y,x,dy,dx,delay,s,date)

head(data)


y <- CoFESSNARX::returnTARGET(data)
x <- CoFESSNARX::returnX(data,F)
#x <- CoFESSNARX::returnX(data,F)
print(head(y))
print(head(x))
```
